Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-28 paper.qmd
Update paper.qmd

[Python]{.proglang} packages, however, require special handling. Unlike CRAN or
the Julia package registry, PyPI (the [Python]{.proglang} Package Index)
contains over 500,000 packages with highly variable quality—many lack proper
build specifications, have broken dependencies, or are abandoned. Because of
this heterogeneity, [Nix]{.pkg} maintainers cannot automatically package all of
PyPI as they do for CRAN or the [Julia]{.proglang} registry. Instead, they
manually curate and package individual [Python]{.proglang} packages that meet
quality standards. This means some [Python]{.proglang} packages, or specific
versions, may not be available through [Nix]{.pkg}.

@@

| Dimension | Traditional Stack | Nix-Based Framework |
|-----------|------------------|---------------------|
| **Paradigm** | Imperative composition of specialized tools | Declarative specification in unified framework |
| **Environment setup** | Step-by-step shell commands | High-level specification |
| **Workflow definition** | File-based rules in Make syntax | Object-based graph in R syntax |
| **Scientific code** | Wrapped in CLI parsing and file I/O boilerplate | Pure functions with no plumbing |
| **File count** | 9 files (4 for orchestration and plumbing) | 6 files (2 for orchestration) |
| **Expertise required** | Linux system administration, Make syntax | R programming |
| **Spatial reproducibility** | Strong | Strong |
| **Temporal reproducibility** | Moderate (vulnerable to drift) | Strong |

>>>>

[Python]{.proglang} packages, however, require special handling. Unlike CRAN or
the Julia package registry, PyPI (the [Python]{.proglang} Package Index)
contains over 500,000 packages with highly variable quality—many lack proper
build specifications, have broken dependencies, or are abandoned. Because of
this heterogeneity, [Nix]{.pkg} maintainers cannot automatically package all of
PyPI as they do for CRAN or the [Julia]{.proglang} registry. Instead, they
manually curate and package individual [Python]{.proglang} packages that meet
quality standards.^[On October 2025, this amounts to roughly 10000 packages, 
according to the search engine for [Nix]{.pkg} packages.] 
This means some [Python]{.proglang} packages, or specific
versions, may not be available through [Nix]{.pkg}.

@@

| Dimension | Traditional Stack | Nix-Based Framework with rix and rixpress |
|-----------|------------------|---------------------|
| **Paradigm** | Imperative composition of specialized tools | Declarative specification in unified framework |
| **Environment setup** | Step-by-step shell commands | High-level specification |
| **Workflow definition** | File-based rules in Make syntax | Object-based graph in R syntax |
| **Scientific code** | Wrapped in CLI parsing and file I/O boilerplate | Pure functions with no plumbing |
| **File count** | 9 files (4 for orchestration and plumbing) | 6 files (2 for orchestration) |
| **Expertise required** | Linux system administration, Make syntax | R programming |
| **Spatial reproducibility** | Strong | Strong |
| **Temporal reproducibility** | Moderate (vulnerable to drift) | Strong |
